Gut-resident microorganisms and their genes are associated with cognition and neuroanatomy in children

### 2023-03-31 manuscript/main.tex
change title
vanja comments 1

\title{Gut-resident microbial species and their genes are associated with cognition and neuroanatomy in children}

@@

The gut and the brain are intimately linked. Signals from the brain
reach the gut through the autonomic nervous system and the endocrine
system, and the gut can communicate with the brain through the vagus
nerve and through endocrine and immune (cytokine) signaling molecules
\cite{cerdoEarlyNutritionGut2019,pronovostPerinatalInteractionsMicrobiome2019,sharonCentralNervousSystem2016,togniniGutMicrobiotaPotential2017}.
In addition, the products of microbial metabolism generated in the gut can
influence the brain, both indirectly by stimulating the enteric nervous
and immune systems, as well as directly through molecules that enter
circulation and cross the blood-brain barrier. Causal links between the
gut microbiome and neural development, particularly atypical
development, are increasingly being identified
\cite{spichakMiningMicrobesMental2021}.
Both human epidemiology and animal models point to the effects
of gut microbes on the development of autism spectrum disorder
\cite{laueProspectiveAssociationsInfant2020,wanUnderdevelopmentGutMicrobiota2021},
and specific microbial taxa have been associated with depression
\cite{mayneris-perxachsMicrobiotaAlterationsProline2022,valles-colomerNeuroactivePotentialHuman2019}
and Alzheimer's disease
\cite{fungInteractionsMicrobiotaImmune2017,kimProbioticSupplementationImproves2021}.
But information about this "microbiome-gut-brain
axis" in normal neurocognitive development remains lacking,
particularly early in life.

The first years of life are critical developmental windows for both the
microbiome and the brain
\cite{laueDevelopingMicrobiomeBirth2022}.
Fetal development is believed to be mostly sterile, but newborns
are rapidly seeded at birth through contact with the birth canal (if
birthed vaginally), caregivers, food (breast milk or formula),
and other environmental sources
\cite{backhedDynamicsStabilizationHuman2015,bokulichAntibioticsBirthMode2016}.
The early microbiome is characterized by low microbial
diversity, rapid succession and evolution, and domination by
Actinobacteria, particularly the genus \emph{Bifidobacterium},
Bacteroidetes, especially \emph{Bacteroides}, and Proteobacteria
\cite{koenigSuccessionMicrobialConsortia2011}.
Many of these bacteria have specialized metabolic capabilities
for digesting human breast milk, such as \emph{Bifidobacterium longum}
subsp. \textit{infantis} and \emph{Bacteroides fragilis}
\cite{selaGenomeSequenceBifidobacterium2008}.
Upon the introduction of solid foods, the gut
microbiome undergoes another categorical transformation;
its diversity increases, and most taxa of the infant microbiome
are replaced by taxa more reminiscent of adult microbiomes
\cite{backhedDynamicsStabilizationHuman2015}.
Prior studies have typically focused on either infant microbiomes or
adult microbiomes, since performing statistical analyses across this
transition poses particular challenges. Nevertheless, since this
transition coincides with critical neural developmental windows and
associated neurodevelopmental processes including myelination, neurogenesis, and synaptic pruning,
investigation across this solid-food boundary is important
\cite{tauNormalDevelopmentBrain2010}.

@@

To begin to address this need, we investigated trends and associations between
the gut microbiome and
neurocognitive development in a large cohort of healthy and neurotypically-developing
children from infancy through 10 years of age. Gut microbial
communities were assessed using shotgun metagenomic sequencing, enabling
profiling at both the taxonomic and gene-functional level.
Cognitive skills and abilities were measured using age-appropriate
psychometric assessments of cognitive function s, i.e.,
the Mullen Scales of Early Learning (MSEL) \cite{mullenMullenScalesEarly1995},
Weschler Preschool and Primary Scale of Intelligence, 4th Edition (WPPSI-4) 
\cite{wechslerWechslerPreschoolPrimary2012}
and Weschler Intelligence Scale for Children, 5th Edition (WISC-V)
\cite{wechslerWechslerIntelligenceScale1949}.
Finally, we also assessed emerging brain structure using magnetic resonance imaging
(MRI). Through a combination of classical statistical analysis and machine
learning, we find that the development of the gut microbiome,
children's cognitive abilities, and brain structure are intimately linked, with
both microbial taxa and gene functions able to predict cognitive
performance and brain structure.

>>>>

\title{Gut-resident microorganisms and their genes are associated with cognition and neuroanatomy in children}

@@

The gut and the brain are intimately linked. Signals from the brain
reach the gut through the autonomic nervous system and the endocrine
system, and the gut can communicate with the brain through the vagus
nerve and through endocrine and immune (cytokine) signaling molecules
\cite{cerdoEarlyNutritionGut2019,pronovostPerinatalInteractionsMicrobiome2019,sharonCentralNervousSystem2016,togniniGutMicrobiotaPotential2017}.
In addition, the products of microbial metabolism generated in the gut can
influence the brain, both indirectly by stimulating the enteric nervous
and immune systems, as well as directly through molecules that enter
circulation and cross the blood-brain barrier. Causal links between the
gut microbiome and neural development, particularly atypical
development, are increasingly being identified
\cite{spichakMiningMicrobesMental2021}.
Both human epidemiology and animal models point to the effects
of gut microbes on the development of autism spectrum disorder
\cite{laueProspectiveAssociationsInfant2020,wanUnderdevelopmentGutMicrobiota2021},
and specific microbial taxa have been associated with depression
\cite{mayneris-perxachsMicrobiotaAlterationsProline2022,valles-colomerNeuroactivePotentialHuman2019}
and Alzheimer's disease
\cite{fungInteractionsMicrobiotaImmune2017,kimProbioticSupplementationImproves2021}.
However, information about this "microbiome-gut-brain
axis" in normal neurocognitive development remains lacking,
particularly early in life.

The first years of life are critical developmental windows for both the
microbiome and the brain
\cite{laueDevelopingMicrobiomeBirth2022}.
Fetal development is believed to occur in a sterile environment, but newborns
are rapidly seeded at birth through contact with the birth canal (if
birthed vaginally), caregivers, food (breast milk or formula),
and other environmental sources
\cite{backhedDynamicsStabilizationHuman2015,bokulichAntibioticsBirthMode2016}.
The early microbiome is characterized by low microbial
diversity, rapid succession and evolution, and domination by
Actinobacteria, particularly the genus \emph{Bifidobacterium},
Bacteroidetes, especially \emph{Bacteroides}, and Proteobacteria
\cite{koenigSuccessionMicrobialConsortia2011}.
Many of these bacteria have specialized metabolic capabilities
for digesting human breast milk, such as \emph{Bifidobacterium longum}
subsp. \textit{infantis} and \emph{Bacteroides fragilis}
\cite{selaGenomeSequenceBifidobacterium2008}.
Upon the introduction of solid foods, the gut
microbiome undergoes another categorical transformation;
its diversity increases, and most taxa of the infant microbiome
are replaced by taxa more reminiscent of adult microbiomes
\cite{backhedDynamicsStabilizationHuman2015}.
Prior studies have typically focused on either infant microbiomes or
adult microbiomes, since performing statistical analyses across this
transition poses particular challenges. Nevertheless, since this
transition coincides with critical neural developmental windows and
associated neurodevelopmental processes including myelination, neurogenesis, and synaptic pruning,
investigation across this solid-food boundary is important
\cite{tauNormalDevelopmentBrain2010}.

@@

To begin to address this need, we investigated relationship between
the gut microbiome and
neurocognitive development in a large cohort of healthy and neurotypically-developing
children from infancy through 10 years of age. Gut microbial
communities were assessed using shotgun metagenomic sequencing, enabling
profiling at both the taxonomic and gene-functional level.
Cognitive skills and abilities were measured using age-appropriate
psychometric assessments of cognitive function, i.e.,
the Mullen Scales of Early Learning (MSEL) \cite{mullenMullenScalesEarly1995},
Weschler Preschool and Primary Scale of Intelligence, 4th Edition (WPPSI-4) 
\cite{wechslerWechslerPreschoolPrimary2012}
and Weschler Intelligence Scale for Children, 5th Edition (WISC-V)
\cite{wechslerWechslerIntelligenceScale1949}.
Finally, we also assessed emerging brain structure using magnetic resonance imaging
(MRI). Through a combination of classical statistical analysis and machine
learning, we find that the development of the gut microbiome,
children's cognitive abilities, and brain structure are intimately linked, with
both microbial taxa and gene functions able to predict cognitive
performance and brain structure.
